---
title: "sta426-week1-ex"
author: "Dipayan Akhuli"
format: 
  html:
    code-fold: true
    code-tools: true
editor: visual
---

```{r}
#| label: load-packages
#| include: false

library(ggplot2)
library(scales)
```

## Week 1 Exercise Part B

In this exercise, 100 values are sampled from a log-normal distribution and the data is plotted as a histogram and as a violin plot, both on linear and log scales.

## Sampling values and calculating statistics

`vals` stores $n = 100$ values sampled from a log-normal distribution with parameters $\mu = 5$ and $\sigma = 0.5$.

```{r}

vals <- data.frame(x = rlnorm(n = 100, meanlog = 5, sdlog = 0.5))

```

The theoretical mean and standard deviation of the log-normal distribution are given by:

$\mu^{*} = e^{\left(\mu + \frac{{\sigma}^2}{2}\right)} \approx 168$

$\sigma^{*} = \sqrt{e^{2\mu + {\sigma}^2} (e^{{\sigma}^2} - 1)} \approx 90$

Here, the observed mean and standard deviation of the sampled values are:

$\mu^{*}_{obs} =$ `r round(mean(vals$x), digits = 2)`

$\sigma^{*}_{obs} =$ `r round(sd(vals$x), digits = 2)`

The observed statistics are indeed quite close to their theoretical values!

## Histograms

The 100 sampled values stored in `vals` are plotted as a histogram on *(a)* a linear scale, and *(b)* a log scale.

```{r}
#| label: fig-hist
#| fig-cap: "Histograms of sampled data"
#| fig-subcap: 
#|     - "Linear x-axis"
#|     - "Log x-axis"
#| layout-ncol: 2
#| column: page

ggplot(vals, aes(x)) +
  geom_histogram(binwidth = 30, fill = "#62b6cb", color = "black") +
  scale_x_continuous(expand = c(0, 0), breaks = seq(0, 1000, by = 100)) +
  scale_y_continuous(expand = c(0, 0), breaks = seq(0, 100, by = 4)) +
  labs(x = "Sampled values", y = "Frequency") +
  theme_bw() +
  theme(
    axis.title.y = element_text(vjust = +3),
    axis.title.x = element_text(vjust = -1),
    panel.border = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5, lineend = "square")
  )

ggplot(vals, aes(x)) +
  geom_histogram(binwidth = 0.15, fill = "#ffbf69", color = "black") +
  scale_x_continuous(
    expand = c(0, 0),
    trans = log_trans(),
    breaks = trans_breaks("log", function(x) exp(x)),
    labels = trans_format("log", math_format(e^.x))
  ) +
  scale_y_continuous(expand = c(0, 0), breaks = seq(0, 100, by = 4)) +
  labs(x = "log(Sampled values)", y = "Frequency") +
  theme_bw() +
  theme(
    axis.title.y = element_text(vjust = +3),
    axis.title.x = element_text(vjust = -1),
    panel.border = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5, lineend = "square")
  )

```

The samples from the log-normal distribution are right-skewed when plotted on a linear scale (Figure 1 *(a)*). Upon performing a log transform, they follow a normal distribution which is now reflected in the symmetry of the histogram about the mean of the samples (Figure 1 *(b)*).

## Violin plot

The 100 sampled values stored in `vals` are now plotted as a violin plot, again on *(a)* a linear scale, and *(b)* a log scale, with an overlaid box plot showing the median and the first and third quartiles.

```{r}
#| label: fig-viol
#| fig-cap: "Violin plots of sampled data"
#| fig-subcap: 
#|     - "Linear y-axis"
#|     - "Log y-axis"
#| layout-ncol: 2
#| column: page

ggplot(vals, aes(x = rep(1, 100), y = x)) + 
  geom_violin(fill = "#62b6cb", color = "#219ebc") +
  geom_boxplot(width = .1, fill = "#219ebc", outlier.colour = NA) +
  stat_summary(fun = median, geom = "point", fill = "white", shape = 21, size = 2.5) +
  labs(x = NULL, y = "Sampled values") +
  theme_bw() +
  theme(
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    axis.title.y = element_text(vjust = +3),
    panel.border = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5, lineend = "square")
  )

ggplot(vals, aes(x = rep(1, 100), y = x)) + 
  geom_violin(fill = "#ffbf69", color = "#fca311") +
  geom_boxplot(width = .1, fill = "#fca311", outlier.colour = NA) +
  stat_summary(fun = median, geom = "point", fill = "white", shape = 21, size = 2.5) +
  scale_y_continuous(
    expand = c(0, 0),
    trans = log_trans(),
    breaks = trans_breaks("log", function(x) exp(x)),
    labels = trans_format("log", math_format(e^.x))
  ) +
  labs(x = NULL, y = "log(Sampled values)") +
  theme_bw() +
  theme(
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    axis.title.y = element_text(vjust = +3),
    panel.border = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5, lineend = "square")
  )

```

As expected from the histogram in Figure 1 *(a)*, the "violin" in Figure 2 *(a)* is bottom-heavy, which similarly becomes symmetric about the mean upon performing a log transform (Figure 2 *(b)*).
