---
title: "Exercise 09"
author: "Dipayan Akhuli (dipayan-akhuli)"
date: today
format: 
    html:
      toc: true
      self-contained: true
      highlight-style: github
      code-line-numbers: true
editor_options: 
    chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, output=FALSE}
library(ggplot2)
```

## Import results

We first import the `edgeR_dge.rds` object, and then identify the top 10 genes whose mean logCPM expression values differ the most between CHIRON-treated and untreated samples.

```{r }
dge <- readRDS("/Users/dipayan/scratch/ARMOR/sta426_dataset/output/outputR/edgeR_dge.rds")

# identify top 10 differentially expressed genes
top10 <- order(abs(rowMeans(dge[["sg"]]@assays@data@listData[["logcpm"]][,1:3]) - rowMeans(dge[["sg"]]@assays@data@listData[["logcpm"]][,4:6])), decreasing=TRUE)[1:10]
dfTop10 <- data.frame(t(dge[["sg"]]@assays@data@listData[["logcpm"]][top10,]), condition=rep(c("CHIRON-treated", "WT unstimulated"), each=3))
rownames(dge[["sg"]]@assays@data@listData[["logcpm"]])[top10]
```

We now make a violin plot of the logCPM expression values of the gene THOC5 across all 6 samples grouped by treatment.

```{r }
#| label: viol-logCPM
ggplot(dfTop10, aes(x=condition, y=dfTop10[,2])) + 
  geom_violin() +
  geom_boxplot(width=.05, outlier.colour=NA, position=position_dodge(width=0.9)) +
  labs(title=colnames(dfTop10)[2], x="Samples", y="logCPM expression values") +
  theme_bw() +
  theme(axis.title.x=element_text(vjust=-1), axis.title.y=element_text(vjust=+3), panel.border=element_blank(), axis.line=element_line(color="black", linewidth=0.5, lineend="square"))
```

We see that the two distributions are quite different, thus, THOC5 is differentially expressed upon CHIRON treatment as compared to the unstimulated WT samples. However, the mean values of the two distributions are quite close, so the extent of differential expression is not that high.

..
